PMV formulations: which one is the most accurate?

### 2022-07-04 Manuscript/src/sections/introduction.tex
paper: revised intro, methodology

To overcome this problem, the \gls{55} Standard uses a modified version of the original model developed by Fanger, the \gls{pmv-ce}~\cite{ashrae552020}.
The process used by the Standard is summarized in the flowchart depicted in Figure~\ref{fig:flowchart_pmv_ce}.
For \ac{v} higher than \qty{0.1}{\m\per\s} the \gls{55} prescribes the use of the \ac{ce}.
The value of \ac{ce}, which is calculated using the \ac{set} equation, is subtracted from both the value of \ac{tdb} and \ac{tr}.
The resulting values become the new inputs into the \ac{pmv} model.
The \ac{ce} accounts for convective heat losses from the person to the environment, thus, the value of \ac{v}, used to calculate the \ac{pmv}, is set to \qty{0.1}{\m\per\s}.
While the other three input parameters (\ac{clo}, \ac{met}, and \ac{rh}) remain unchanged.
Consequently, for a given thermal environment, results from the two \ac{pmv} formulation differ when the value of \ac{v} is higher than \qty{0.1}{\m\per\s}.

@@

This is a source of confusion for many users worldwide since both the \gls{7730} and the \gls{55} are the widely adopted and incorporated in many national and international building codes worldwide.
To further complicate this issue, over the course of the last \num{50} years, several other formulations of the \ac{pmv} model have been proposed.
Among the most notable formulations there are the: \gls{pmvs}, \gls{pmvg}, \gls{epmv}, \gls{apmv}, and \ac{athb}.

% todo pmv SET
% todo pmv Gagge

@@

To our knowledge, no other study compared the accuracy all the aforementioned aforementioned PMV models.
\mycite{Cheung2019} determined the accuracy of the \ac{pmv} model, by comparing its results with the \ac{tsv} from the ASHRAE ASHRAE Global Thermal Comfort Database II.
While \textbf{COBEE paper} compared the accuracy of the \gls{pmv-ce} and the \ac{pmv} models.

This study aims to specifically fill this knowledge gap by comparing the results of the \ac{pmv} formulations developed over the course of the last years with self-reported \ac{tsv} contained in the ASHRAE Global Thermal Comfort Database II v2.1, referred to as the \gls{db2} thereafter.

>>>>

To overcome this problem, the \gls{55} Standard uses a modified version of the original model developed by Fanger, the \gls{pmv-ce}~\cite{ashrae552020}.
The process used to calculate the \gls{pmv-ce} is summarized in the flowchart depicted in Figure~\ref{fig:flowchart_pmv_ce}.
For \ac{v} higher than \qty{0.1}{\m\per\s} the \gls{55} prescribes the use of the \ac{ce}.
The value of \ac{ce}, which is calculated using the \ac{set} equation, is subtracted from both the value of \ac{tdb} and \ac{tr}.
The resulting values become the new inputs into the \ac{pmv} model.
The \ac{ce} accounts for convective heat losses from the person to the environment, thus, the value of \ac{v}, used to calculate the \ac{pmv}, is set to \qty{0.1}{\m\per\s}.
While the other three input parameters (\ac{clo}, \ac{met}, and \ac{rh}) remain unchanged.
Consequently, for a given thermal environment, results from the two \ac{pmv} formulation differ when the value of \ac{v} is higher than \qty{0.1}{\m\per\s}.

@@

This is a source of confusion for many users worldwide since both the \gls{7730} and the \gls{55} are the widely adopted and incorporated in many national and international building codes worldwide.
To further complicate this issue, over the course of the last decades, several other formulations of the \ac{pmv} model have been proposed.
Among the most notable formulations there are the: \gls{pmvs}, \gls{pmvg}, \gls{epmv}, \gls{apmv}, and \ac{athb}.

% todo describe pmv SET
% todo describe pmv Gagge

@@

To our knowledge, no other study compared the accuracy of all the aforementioned PMV models, aside from a conference paper we previously published that stands at the basis of this work.
In particular, no previous study compared the accuracy of the \ac{pmv} and \ac{pmv-ce} models which are currently used in the \gls{7730} and \gls{55} Standards.
Previous works were limited to the determination of the accuracy of the \ac{pmv} model.
For example, some notable works include, the work from \mycite{Doherty1988} who evaluated the ability of the \ac{pmv} and \ac{set} models to predict several physiological variables (i.e., skin temperature, core temperature, and skin wettedness) under a wide range of still air environments and metabolic rates.
They concluded that the \ac{pmv} model is accurate for simulations of resting subjects but its accuracy decreases as a function of metabolic rate.
Humphreys and Nicol estimated the effects of measurement and formulation error on predicting thermal sensation using the \ac{pmv}~\cite{Humphreys2000}.
They used the ASHRAE Thermal Comfort database v1.0 and determined that the measurement error and the error introduced by the \ac{pmv} model formulation had a similar and non-negligible contribution.
Humphreys and Nicol also determined the validity of the \ac{pmv} for predicting comfort votes collected in field studies~\cite{Humphreys2002}.
They concluded that the \ac{pmv} range of applicability should be significantly reduced and it fails to predict the extent of thermal dissatisfaction of people in buildings.
The \ac{pmv} was free from bias only when it was used to predict thermal neutrality.
While, \mycite{Cheung2019} determined the accuracy of the \ac{pmv} model, by comparing its results with the \ac{tsv} from the ASHRAE Global Thermal Comfort Database II.

This study aims to specifically fill this knowledge gap by comparing the results of the major \ac{pmv} formulations developed over the course of the last decades with self-reported \ac{tsv} contained in the ASHRAE Global Thermal Comfort Database II v2.1, referred to as the \gls{db2} thereafter.
Our results have significant and profound implications for both researchers and practitioners since it provides quantitative evidence on which formulation of the \ac{pmv} model is more accurate.

### 2022-07-04 Manuscript/src/sections/methodology.tex
paper: revised intro, methodology

The \gls{db2} has \var{all_entries_db} independent recordings from various field investigations throughout the world and it is the largest of its kind.
\Ac{tsv} were collected using a \gls{rhrn} survey and environmental parameters were measured and logged while participants completed their survey.
The Comfort DB contains data collected by researchers in cross-sectional field studies that have been published in peer reviewed journals.
%We compared the results obtained by the \ac{pmv} and the \gls{pmv-ce} with the self-reported \ac{tsv}.
%    All the results were calculated using the function pmv_ppd [6] included in pythermalcomfort v1.10.0 a Python package for thermal comfort calculations.

\subsection{Data Processing and Cleaning}\label{subsec:data-processing-and-cleaning}
Not all researchers logged all the six parameters required to calculate the PMV.\@
We, therefore, decided to remove all entries that did not have any of the following variables: \ac{tdb}, \ac{rh}, \ac{v}, \ac{clo}, and \ac{met}.
These constraints filtered out approximately \qty{28}{\percent} of the data in the Comfort DB.\@  % todo check the above value
We decided to keep only entries that did not have \ac{tr} data but instead had the \ac{to} value.
Since \ac{tr} is a required input in the \ac{pmv} model, it was calculated using the following equation: 2\ac{to} – \ac{tdb}.
Alternatively, if not value of \ac{tr} was available we assumed \ac{tr} equal to \ac{tdb}.
We are aware that these assumptions may introduce an error, but removing samples without \ac{tr} would cause an additional \qty{44}{\percent} drop in the total entries in the database and there is evidence that in many conditions is possible to assume assumed \ac{tr} equal to tdb [7].  % todo check the above value and fix citation

Both the \gls{55} and \gls{7730} specify a set of applicability limits, we hence filtered out those data points which did not meet the inclusion criteria of both Standards.
The rationale is that the model accuracy should only be tested in within its applicability limits.
The limits we used to filter the data are as follows: \qty{10}{\celsius} $\leq$ \ac{tdb} $\leq$ \qty{30}{\celsius}, \qty{10}{\celsius} $\leq$ \ac{tr} $\leq$ \qty{40}{\celsius}, \qty{0}{\m\per\s} $\leq$ \ac{v} $\leq$ \qty{1}{\m\per\s}, \qty{0}{clo} $\leq$ \ac{clo} $\leq$ \qty{1.5}{clo}, 0 Pa $\leq$ water vapour partial pressure $\leq$ 2000 Pa, and \qty{1}{met} $\leq$ \ac{met} $\leq$ \qty{4}{met}.
We then calculated the adjusted total clothing insulation and relative airspeed as required by both Standards.
We used these inputs to calculate the \ac{pmv} values.  % todo check with marcel if this is required in his model too

The \ac{pmv} model is only applicable when its absolute value is lower than 2 \cite{Fanger1970, iso7730}.
However, since the thermal sensation was measured with at least a seven-point scale we decided also to show entries with |\ac{tsv}| or |\ac{pmv}| $\leq$ than \num{3.5}.
We used Python to analyse and plot the results.

@@

>>>>

The \gls{db2} has \var{entries_db_all} independent recordings from various field investigations throughout the world and it is the largest of its kind.
\Ac{tsv} were collected using a \gls{rhrn} survey and environmental parameters were measured and logged while participants completed their survey.
The Comfort DB contains data collected by researchers in cross-sectional field studies that have been published in peer reviewed journals.
% We compared the results obtained by the \ac{pmv} and the \gls{pmv-ce} with the self-reported \ac{tsv}.
%    All the results were calculated using the function pmv_ppd [6] included in pythermalcomfort v1.10.0 a Python package for thermal comfort calculations.

\subsection{Data Preparation and Cleaning}\label{subsec:data-processing-and-cleaning}
Not all researchers logged all the six parameters required to calculate the PMV.\@
We, therefore, decided to remove all entries that did not have any of the following variables: \ac{tdb}, \ac{rh}, \ac{v}, \ac{clo}, \ac{met}, and \ac{tsv}.
These constraints filtered out approximately \var{entries_db_valid} of the data in the Comfort DB.\@
We decided to keep entries that did not have \ac{tr}.
Since \ac{tr} is a required input in the \ac{pmv} model, it was calculated using the following equation: 2\ac{to} – \ac{tdb}.
Alternatively, if \ac{to} was not available we assumed \ac{tr} equal to \ac{tdb}.
We are aware that these assumptions may introduce an error, but removing samples without \ac{tr} would cause an additional \var{entries_db_valid_no_tr} drop in the data used to compare the models.
There is also evidence that in many conditions is possible to assume \ac{tr} equal to \ac{tdb} without committing a major error~\cite{Dawe2020}.

Both the \gls{55} and \gls{7730} specify a set of applicability limits, we hence filtered out those data points which did not meet the inclusion criteria of both Standards.
The rationale is that the model accuracy should only be tested in within its applicability limits.
The limits we used are as follows: \qty{10}{\celsius} $\leq$ \ac{tdb} $\leq$ \qty{30}{\celsius}, \qty{10}{\celsius} $\leq$ \ac{tr} $\leq$ \qty{40}{\celsius}, \qty{0}{\m\per\s} $\leq$ \ac{v} $\leq$ \qty{1}{\m\per\s}, \qty{0}{clo} $\leq$ \ac{clo} $\leq$ \qty{1.5}{clo}, 0 Pa $\leq$ water vapour partial pressure $\leq$ 2000 Pa, and \qty{1}{met} $\leq$ \ac{met} $\leq$ \qty{4}{met}.
We then calculated the adjusted total clothing insulation and relative airspeed as required by both Standards.
We used these inputs to calculate the \ac{pmv} values.  % todo check with marcel if this is required in his model too

Fanger and the \gls{7730} state that the \ac{pmv} should only be used when its absolute value is lower than 2~\cite{Fanger1970, iso7730}.
However, since the thermal sensation was measured with at least a seven-point scale we decided also to show entries with $|$\ac{tsv}$|$ or $|$\ac{pmv}$|$ $\leq$ than \num{3.5}.

The values of \ac{pmv} and \gls{pmv-ce} are available inside the \gls{db2}.
We calculated the values of \gls{pmvs}, \gls{pmvg}, and \ac{athb} using pythermalcomfort v2.5.1~\cite{Tartarini2020a}.
We used Python to analyse and plot the results.

@@

\subsection{Model Validation}\label{subsec:model-validation}
The \ac{pmv} model was developed to predict the average thermal sensation of a large group of occupants sharing the same environment.
While the \gls{db2} contains data from a sufficiently wide population no information is available on where exactly these data were collected.
We only have access to building level data and no information about weather a group of occupants was sharing the same thermal environment.
We, therefore, decided to first group the participants responses by their \ac{tsv} and we determined how many of these votes were correctly labelled by the \ac{pmv} model.
This approach is useful to perform an overall assessment of the accuracy of the model, but it introduces an error due to the rounding of the \ac{pmv} models output and \ac{tsv} values.
The latter were sometimes collected using a continuous scale and not a discrete one, in other words they could take any value in the range of \numrange{-3.0}{3.0}.
To remove this rounding error, we then went on subtracted the \ac{tsv} value from the respective \ac{pmv} value.
These deltas quantify the success of the model in predicting \ac{tsv}.
However, on their own are a low precision estimate of the overall accuracy of the model~\cite{Humphreys2002}.
These values were then binned using several independent variables (e.g., \ac{tdb}, \ac{rh}).
If the \ac{pmv} formulation is bias-free, the distribution of any batch derived from these differences would have a mean value that was little different from zero.
The standard deviation would reflect the combined effect of the people's individual differences as well as any errors in the \ac{pmv} formulation~\cite{Humphreys2002}.
According to \mycite{Humphreys2002} the model can be considered accurate if the above-mentioned deltas are within \numrange{-0.25}{0.25}.
% todo why not 0.5 since the delta between a PMV of 2.5 and a self-reported vote of either 2 or 3 (closest points to 2.5) is 0.5?

\subsection{Sources of Error}\label{subsec:sources-of-error}
One minor but not negligible source of error is the use of a discrete scale to assess thermal sensation.
Most of the \ac{tsv} in the \gls{db2} were collected using a discrete scale while the \ac{pmv} output is continuous.
For example, even if the model was 100~\% accurate and the estimated \ac{pmv} values was \num{2.5} the person could only report to be `warm (+2)' or `hot (+3)'.

Measurement errors are also inevitably present in the dataset.
These will add a random error to the model output.

Errors are also caused by the fact that the \ac{pmv} model is an approximation of a highly complex system (the human body and mind) which interacts with the surrounding environment.
Fanger could not, therefore, include all the factors that contribute to the comfort vote and some of them have been omitted.
Finally, it should be noted that the \ac{pmv} model has been developed considering based on the assumption of steady-state heat transfer, however, this never precisely occurs since the human thermo regulatory system is always actively engaged to ensure a stable core temperature.
However, as \mycite{Humphreys2002} note, these approximations and errors present in the \ac{pmv} model, if the model is accurate, should combine to produce a quasi-random error in the prediction.
This would increase the standard deviation of the delta between the \ac{tsv} and \ac{pmv}, but if the model is sound, they should produce a negligible bias~\cite{Humphreys2002}.
On the other hand, if the model is neither accurate nor representative of the real system a significant bias will arise.

### 2022-07-04 Manuscript/src/sections/results.tex
paper: revised intro, methodology

A total of \var{tot_surveys} entries met the inclusion criteria listed in the Methodology Section.
The distribution of the six input variables used to calculate the PMV indices is depicted in Figure~\ref{fig:dist_input_data}.

@@

>>>>

\subsection{Comfort DB overview}\label{subsec:comfort-db-overview}

A total of \var{entries_db_used} entries met the inclusion criteria listed in the Methodology Section.
The distribution of the six input variables used to calculate the PMV indices is depicted in Figure~\ref{fig:dist_input_data}.

@@

Figure~\ref{fig:dist_other_data} shows the distribution of the age, height, weight, and running mean outdoor temperature grouped by sex.
Less than half (\num{23300}) of the total entries had information about the participant's sex.
Out of those data were almost equally distributed among male \qty{52}{\percent} and females.
Ages are not normally distributed, the same is true for the running mean outdoor temperature.
About half of the participants (\qty{52}{\percent}) were aged between \num{20} and \num{35} years old.
While \qty{56}{\percent} of the running mean outdoor temperature values where between \qtyrange{10}{25}{\celsius}.
Mens were significantly heavier and taller than females.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/dist_other_data}
    \caption{Distribution of age, height, weight, and running mean outdoor temperature.
    The data are grouped by sex.}
    \label{fig:dist_other_data}
\end{figure}

Figure~\ref{fig:bar_plot_tp_by_ts} shows on the left the percentage of thermal preference votes grouped by each \ac{tsv}.
The left Figure also reports the number of points for each \ac{tsv}.
Only \var{entries_with_tp} that had information about the thermal preference.
Thus, on the right we show a bar chart depicting the distribution of all the \ac{tsv} votes, we have used in the analysis.
The \gls{db2} in respect to the \ac{tsv} is unbalanced.
Approximately \var{perc_tsv_neutral} of all the entries have a \ac{tsv} of `Neutral'.
While less than \var{perc_tsv_hot} of the total sample of participants reported to be either `Hot' or `Cold'.
In the \gls{db2} more than \qty{50}{\percent} of participants who were either `slightly warm' or `slightly cool' wanted to be `cooler' or `warmer', respectively.
However, in thermal comfort research it is generally assumed that people who are `slightly warm' or `slightly cool' are thermally comfortable.
This finding challenges the above assumption.
Moreover, these results provide additional evidence that the use of the thermal preference scale is better suited to determine how people perceive their thermal environment.
The thermal sensation vote clearly allows to determine whether participants would take action or not to modify their current thermal environment.
The \ac{tsv}, on the contrary, does not allow to determine which actions should be taken to increase thermal comfort conditions for occupants.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/bar_plot_tp_by_ts}
    \caption{The left Figure shows the percentage of thermal preference votes for each thermal sensation vote.
    The numbers on the right side of each bar show the number of points for each \ac{tsv}.
    The right figure shows the total number of data points grouped by \ac{tsv}.
    Each bin in the Figure of the right has more data points than in the Figure on the left since information about the occupants thermal preference were not always available.}
    \label{fig:bar_plot_tp_by_ts}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/bubble_models_vs_tsv}
    \caption{}
    \label{fig:bubble_models_vs_tsv}
\end{figure}

### 2022-07-04 Manuscript/src/sections/titleAbstract.tex
paper: revised intro, methodology

    \begin{abstract}
        The \ac{pmv} model was originally developed by Fanger to predict the thermal sensation of a large group of occupants sharing the same thermal environment.
        It has seen widespread adoption and it is currently used in the \gls{7730}.
        However, to compensate for the model low prediction accuracy, several other formulations have been proposed, one of which is currently used in the \gls{55}.
        This is a source of confusion for practitioners and researchers who find it difficult to decide which formulation to use.
        We aim to solve this problem by comparing the accuracy of the main \ac{pmv} formulations against \var{tot_surveys} thermal sensation votes collected in buildings.
        % We found that for V>0.1m/s the \gls{pmv-ce} is significantly more accurate (48%) than the PMV (42%) in predicting thermal neutrality.
        % In 95% of the cases, the error was less than one point.
        % The PMV and PMVCE accuracy were 35% and 36% for |PMV|≤2 and 43% and 44% for |PMV|≤1, respectively.
        % Neither PMV model could accurately predict ‘cold’, ‘cool’, ‘warm’, or ‘hot’ thermal sensations.
        % Consequently, the PMV should only be used in conditions close to neutrality within the range |PMV|≤1.
        % We open-sourced the source code.
    \end{abstract}

>>>>

    \begin{abstract}
        The \ac{pmv} model was originally developed by Fanger to predict the thermal sensation of a large group of occupants sharing the same thermal environment.
        It has seen widespread adoption and in its original form it is currently used in the \gls{7730}.
        However, the \ac{pmv} model has a low prediction accuracy and to compensate for that, several other formulations have been proposed, one of which is currently used in the \gls{55}.
        Being the \gls{7730} and \gls{55} the most widely used Standards worldwide, this is a source of confusion for practitioners and researchers who find it difficult to decide which formulation to use.
        We aim to solve this problem by comparing the accuracy of the main \ac{pmv} formulations against \var{entries_db_used} thermal sensation votes collected in buildings.
        % We found that for V>0.1m/s the \gls{pmv-ce} is significantly more accurate (48%) than the PMV (42%) in predicting thermal neutrality.
        % In 95% of the cases, the error was less than one point.
        % The PMV and PMVCE accuracy were 35% and 36% for |PMV|≤2 and 43% and 44% for |PMV|≤1, respectively.
        % Neither PMV model could accurately predict ‘cold’, ‘cool’, ‘warm’, or ‘hot’ thermal sensations.
        % Consequently, the PMV should only be used in conditions close to neutrality within the range |PMV|≤1.
        % We open-sourced the source code.
    \end{abstract}
